---
title: "PA1_template"
author: "Balázs J. Keresztes"
date: "09/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)

options(warn=-1)
```

## Reproducable Research - Course Project 1

The following report is based on the questions provided in the Reproducable Research 
course's first Course Project. It is intended to showcase writing an R Markdown file 
that contains an alaysis on activity monitoring device data using knitr nd the analysis 
itself.


### Loading and preprocessing the data

The data is provided for the assignment. 

*"This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day."*

```{r}
data <- tbl_df(read.csv("activity.csv"))

summary(data)
```

As defined in the assignment, the variables are
```{r}
names(data)
```

where

* **steps**: Number of steps taking in a 5-minute interval (missing values 
are coded as \color{red}{\verb|NA|}NA);
* **date**: The date on which the measurement was taken in YYYY-MM-DD format;
* **interval**: Identifier for the 5-minute interval in which measurement 

was taken.


### What is mean total number of steps taken per day?

Firstly, we're interested in the mean and median steps taken on a daily 
basis. For this, we group the data by days.
```{r}
daily <- data %>%
        group_by(date) %>%
        summarise(total = sum(steps))

head(daily)
```

Now, we extract the number of days in the data; the number of groups in
`daily` that has valid values.
```{r}
n <- nrow(daily[!is.na(daily$total), ])
```

We can determine the mean total steps taken on a single day (ignoring
NA values).
```{r}
mean <- mean(daily$total, na.rm = TRUE)
```

As well as the median value:
```{r}
median <- median(daily$total, na.rm = TRUE)
```

We get the following results:

* ``r paste("n =", n)``
* ``r paste("mean =", mean)``
* ``r paste("median =", median)``

We can check how this number looks like compared in a histogram. The bars represent
a 500 wide interval in which we group the different daily totals, the red vertical
line shows the mean value.
```{r}
qplot(daily$total, geom="histogram", binwidth=500, ylim=c(0,7)) + 
        geom_vline(xintercept = mean, col = "Red", lwd = 1.5) +
        xlab("Daily Total Steps") +
        ylab("Frequency")
```


### What is the average daily activity pattern?

Let's see how the average steps are distributed on a timescale when we take and picture
the intervals of 5 minutes and compare them based on how many steps are taken on average
within them. I'm going to group by intervals, calculate the individual mean values for 
each interval, and plot them. In order to visualise this, I'm also going to convert the 
interval values to something more readable, into hh:mm format.

```{r}
to_hm <- function(x) {
        r <- character(length = length(x))
        for(i in seq_along(x)) {
                if(nchar(x[i]) == 1) { r[i] <- paste0("0:0", x[i]) }
                else if(nchar(x[i]) == 2) { r[i] <- paste0("0:", x[i]) }
                else if(nchar(x[i]) == 3) { r[i] <- paste0(substr(x[i],1,1),":",substr(x[i],2,3)) }
                else { r[i] <- paste0(substr(x[i],1,2),":",substr(x[i],3,4)) } 
        }
        r
}

interval <- data %>% 
        group_by(interval) %>% 
        summarise(mean = mean(steps, na.rm = TRUE))

ggplot(
        data = interval, 
        aes(1:288, mean)) + 
        geom_line() + 
        scale_x_continuous(
                breaks = seq(from=1, to=288, by=36), 
                labels = to_hm(interval$interval[seq(from=1, to=288, by=36)])) +
        xlab("Time (in 5 minute intervals)") +
        ylab("Mean Steps Taken")
```

And as the graph clearly shows the 5-minute interval with the highest average steps taken
can be easily spotted.

```{r}
highest <- interval %>%
        filter(mean == max(mean))
highest
```

This seems to be in line with what we can see on the graph.
